---
title: "Reconstruction: 2024 Global Hunger Index by Severity"
subtitle: "RMIT 050646 — Data Visualisation and Communication, Assignment 2"
author: "Author"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    df_print: paged
    fig_width: 14
    fig_height: 8.4
    fig_caption: true
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 14
    fig_height: 8.4
    fig_caption: true
    keep_tex: false
    latex_engine: xelatex
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  fig.width  = 14,
  fig.height = 8.4,
  dpi        = 220,
  out.width  = "100%"
)
```

# Overview

This R Markdown document reproduces the data visualisation submitted for
RMIT 050646 Assignment 2. It reconstructs the *2024 Global Hunger Index by
Severity* poster map (Welthungerhilfe, Concern Worldwide & IFHV, 2024) with
three design improvements:

1. **Colour and accessibility.** A sequential, colourblind-safe YlOrRd
   palette replaces the original's green-to-red ordinal scale, and the
   single ambiguous grey is split into two distinguishable neutrals
   (high-income excluded vs. genuinely missing data).
2. **Chart-type and information density.** Direct annotations of the
   seven Alarming countries plus an adjacent slope panel showing the
   eight largest reductions in GHI score since 2000 turn a static
   severity snapshot into a chart that answers two related questions at
   once.
3. **Projection and decluttering.** A Robinson projection replaces the
   original's Mercator-style cylindrical projection, a journalistic
   finding-led title replaces the placeholder dataset name, and the
   dense overlay of 100+ country labels is removed.

The data is the **2025 GHI** (Welthungerhilfe et al., 2025), transcribed
from <https://www.globalhungerindex.org/ranking.html>.

# 1. Setup

## 1.1 Required packages

If running this for the first time, install the packages once:

```{r install-packages, eval = FALSE}
install.packages(c(
  "tidyverse", "sf", "rnaturalearth", "rnaturalearthdata",
  "ggrepel", "patchwork", "scales"
))
```

Load them:

```{r load-packages}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(patchwork)
library(scales)
```

## 1.2 Working directory note

Place `ghi_data.csv` in the same folder as this `.Rmd` file. When you
knit, R Markdown automatically uses that folder as the working directory,
so the relative path `"ghi_data.csv"` in the `read_csv()` call below will
resolve correctly.

# 2. Data

## 2.1 Load the 2025 GHI scores

```{r load-data}
ghi <- read_csv(
  "ghi_data.csv",
  show_col_types = FALSE,
  col_types = cols(
    country       = col_character(),
    iso3          = col_character(),
    ghi_2000      = col_double(),
    ghi_2008      = col_double(),
    ghi_2016      = col_double(),
    ghi_2025      = col_double(),
    severity_2025 = col_character(),
    provisional   = col_logical()
  )
)

# Severity is an ordered factor so the legend respects the natural
# progression Low -> Moderate -> Serious -> Alarming
ghi <- ghi %>%
  mutate(severity_2025 = factor(
    severity_2025,
    levels  = c("Low", "Moderate", "Serious", "Alarming"),
    ordered = TRUE
  ))

glimpse(ghi)
```

## 2.2 Quick data summary

```{r data-summary}
ghi %>%
  count(severity_2025, name = "n_countries") %>%
  arrange(severity_2025)
```

# 3. World geometry

Robinson is a pseudo-cylindrical compromise projection used by National
Geographic and the European Commission for world thematic maps. It
distributes area distortion more evenly than Mercator, which inflates
high-latitude regions — a particular problem for hunger data, where the
most-affected continents lie near the equator.

```{r build-world}
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(continent != "Antarctica") %>%
  st_transform(crs = "ESRI:54030")  # Robinson

# High-income countries that the GHI does not score (hunger is below the
# scale's measurement threshold). The original poster fills these with the
# same grey as countries with missing data, conflating two distinct
# epistemic states. We tag them so they get their own neutral fill.
high_income_excluded <- c(
  "USA", "CAN", "GBR", "IRL", "FRA", "DEU", "NLD", "BEL", "LUX", "CHE",
  "AUT", "DNK", "NOR", "SWE", "FIN", "ISL", "ESP", "PRT", "ITA", "GRC",
  "CYP", "MLT", "POL", "CZE", "SVN", "JPN", "KOR", "SGP", "NZL", "AUS",
  "ISR", "TWN", "QAT", "BHR"
)

world_ghi <- world %>%
  left_join(ghi, by = c("iso_a3" = "iso3")) %>%
  mutate(
    display_class = case_when(
      !is.na(severity_2025)            ~ as.character(severity_2025),
      iso_a3 %in% high_income_excluded ~ "High-income (excluded from GHI)",
      TRUE                             ~ "No data"
    ),
    display_class = factor(
      display_class,
      levels = c(
        "Low", "Moderate", "Serious", "Alarming",
        "High-income (excluded from GHI)", "No data"
      )
    )
  )
```

# 4. Colour palette

Sequential YlOrRd ramp validated against deutan, protan, and tritan
colour-vision deficiency simulation. The two non-scored categories are
assigned distinct neutrals — a warm off-white for "high-income excluded"
and a cool grey for "no data" — so they are visually distinguishable.

```{r palette}
ghi_palette <- c(
  "Low"                              = "#fee08b",
  "Moderate"                         = "#fdae61",
  "Serious"                          = "#f46d43",
  "Alarming"                         = "#a50026",
  "High-income (excluded from GHI)"  = "#ebebe0",
  "No data"                          = "#bdbdbd"
)
```

# 5. Annotations: the seven Alarming countries

```{r annotations}
alarming_iso <- c("YEM", "SOM", "SSD", "COD", "BDI", "MDG", "HTI")

alarming_labels <- world_ghi %>%
  filter(iso_a3 %in% alarming_iso) %>%
  mutate(
    label_text = case_when(
      iso_a3 == "YEM" ~ "Yemen\n>=35 (provisional)",
      iso_a3 == "SOM" ~ "Somalia\n42.6",
      iso_a3 == "SSD" ~ "South Sudan\n37.5",
      iso_a3 == "COD" ~ "DR Congo\n37.5",
      iso_a3 == "BDI" ~ "Burundi\n>=35 (provisional)",
      iso_a3 == "MDG" ~ "Madagascar\n35.8",
      iso_a3 == "HTI" ~ "Haiti\n35.7"
    )
  )

# Use point-on-surface for label anchors (centroid can fall outside
# polygons for crescent or fragmented countries)
alarming_labels <- alarming_labels %>%
  mutate(
    geometry_pt = st_point_on_surface(geometry),
    x = st_coordinates(geometry_pt)[, 1],
    y = st_coordinates(geometry_pt)[, 2]
  )
```

# 6. Main map

```{r build-map}
p_map <- ggplot(world_ghi) +
  geom_sf(
    aes(fill = display_class),
    colour    = "white",
    linewidth = 0.18
  ) +
  scale_fill_manual(
    values = ghi_palette,
    name   = "2025 GHI severity",
    labels = c(
      "Low (<= 9.9)",
      "Moderate (10.0 - 19.9)",
      "Serious (20.0 - 34.9)",
      "Alarming (>= 35.0)",
      "High-income -\nnot scored by GHI",
      "No data / insufficient"
    ),
    drop  = FALSE,
    guide = guide_legend(
      title.position = "top",
      title.hjust    = 0,
      ncol           = 1,
      override.aes   = list(colour = "white")
    )
  ) +
  ggrepel::geom_label_repel(
    data               = alarming_labels,
    aes(x = x, y = y, label = label_text),
    size               = 3.0,
    fontface           = "plain",
    label.padding      = unit(0.18, "lines"),
    label.r            = unit(0.10, "lines"),
    label.size         = 0.30,
    colour             = "#1a1a1a",
    fill               = "white",
    segment.colour     = "#555555",
    segment.size       = 0.35,
    box.padding        = 0.55,
    point.padding      = 0.20,
    min.segment.length = 0,
    seed               = 42,
    max.overlaps       = Inf
  ) +
  coord_sf(
    crs    = "ESRI:54030",
    expand = FALSE,
    ylim   = c(-7e6, 8.6e6)
  ) +
  labs(
    title    = "Hunger remains concentrated in Sub-Saharan Africa and conflict-affected states",
    subtitle = paste0(
      "Global Hunger Index 2025 - seven countries are at alarming levels.\n",
      "Six of the seven are in Sub-Saharan Africa or directly affected by armed conflict."
    ),
    caption  = paste0(
      "Data: Welthungerhilfe, Concern Worldwide & IFHV (2025), ",
      "2025 Global Hunger Index - globalhungerindex.org/ranking.html.\n",
      "Boundaries: Natural Earth 1:50m. Projection: Robinson."
    )
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title            = element_text(face = "bold", size = 14,
                                         colour = "#111111",
                                         margin = margin(b = 4)),
    plot.subtitle         = element_text(size = 10.5, colour = "#444444",
                                         lineheight = 1.15,
                                         margin = margin(b = 10)),
    plot.caption          = element_text(size = 8, colour = "#666666",
                                         hjust = 0,
                                         margin = margin(t = 8)),
    plot.caption.position = "plot",
    plot.title.position   = "plot",
    legend.position       = c(0.07, 0.30),
    legend.title          = element_text(face = "bold", size = 9.5),
    legend.text           = element_text(size = 8.6, lineheight = 0.95),
    legend.background     = element_rect(fill = "white", colour = "#dcdcdc",
                                         linewidth = 0.4),
    legend.key.size       = unit(0.45, "cm"),
    legend.spacing.y      = unit(0.20, "cm"),
    plot.background       = element_rect(fill = "white", colour = NA),
    plot.margin           = margin(8, 8, 8, 8)
  )
```

# 7. Side panel: largest GHI reductions, 2000 -> 2025

```{r build-panel}
top_reducers <- ghi %>%
  filter(!is.na(ghi_2000) & !is.na(ghi_2025)) %>%
  mutate(change = ghi_2000 - ghi_2025) %>%
  slice_max(change, n = 8) %>%
  mutate(country = fct_reorder(country, change))

top_reducers
```

```{r panel-plot}
p_panel <- ggplot(top_reducers) +
  geom_segment(
    aes(x = ghi_2025, xend = ghi_2000,
        y = country, yend = country),
    colour    = "#f0d8a8",
    linewidth = 1.3,
    lineend   = "round"
  ) +
  geom_point(
    aes(x = ghi_2000, y = country),
    shape  = 21, size = 3.2,
    fill   = "white", colour = "#a50026", stroke = 1.0
  ) +
  geom_point(
    aes(x = ghi_2025, y = country),
    shape  = 21, size = 3.2,
    fill   = "#fdae61", colour = "#7a3a02", stroke = 0.6
  ) +
  geom_text(
    aes(x = ghi_2000, y = country,
        label = sprintf("%.1f", ghi_2000)),
    nudge_x = 3.0, hjust = 0, size = 2.7, colour = "#a50026"
  ) +
  geom_text(
    aes(x = ghi_2025, y = country,
        label = sprintf("%.1f", ghi_2025)),
    nudge_x  = -3.0, hjust = 1, size = 2.7,
    fontface = "bold", colour = "#7a3a02"
  ) +
  scale_x_continuous(limits = c(-5, 75), expand = c(0, 0)) +
  labs(
    title    = "Largest reductions in hunger",
    subtitle = "GHI score, 2000 vs 2025 - top 8 countries",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold", size = 11,
                                    colour = "#111111"),
    plot.subtitle    = element_text(size = 9, colour = "#555555",
                                    margin = margin(b = 8)),
    axis.text.y      = element_text(size = 9, colour = "#1a1a1a"),
    axis.text.x      = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid       = element_blank(),
    plot.margin      = margin(8, 12, 8, 8),
    plot.background  = element_rect(fill = "white", colour = NA)
  )
```

# 8. Final reconstruction

```{r final-plot, fig.width = 14, fig.height = 8.4}
final_plot <- p_map + p_panel +
  plot_layout(widths = c(3.4, 1.0)) &
  theme(plot.background = element_rect(fill = "white", colour = NA))

final_plot
```

# 9. Save outputs

```{r save-outputs, eval = TRUE}
ggsave(
  "reconstruction.png", final_plot,
  width = 14, height = 8.4, dpi = 300, bg = "white"
)

ggsave(
  "reconstruction.pdf", final_plot,
  width = 14, height = 8.4, bg = "white"
)
```

# References

Besançon, L., Cooper, M., Ynnerman, A., & Vernier, F. (2020). An
evaluation of visualization methods for population statistics based on
choropleth maps. *arXiv preprint arXiv:2005.00324*.
<https://doi.org/10.48550/arXiv.2005.00324>

Brewer, C. A., & Harrower, M. (2013). *ColorBrewer 2.0: Colour advice for
cartography* [Web tool]. The Pennsylvania State University.
<https://colorbrewer2.org/>

Cairo, A. (2019). *How charts lie: Getting smarter about visual
information*. W. W. Norton & Company.

Natural Earth. (2024). *Admin 0 — Countries (1:50m and 1:110m)* [Vector
data set]. Natural Earth. <https://www.naturalearthdata.com/>

Welthungerhilfe, Concern Worldwide, & Institute for International Law of
Peace and Armed Conflict. (2024). *2024 Global Hunger Index: How gender
justice can advance climate resilience and zero hunger*. Welthungerhilfe;
Concern Worldwide; IFHV.
<https://www.globalhungerindex.org/pdf/en/2024.pdf>

Welthungerhilfe, Concern Worldwide, & Institute for International Law of
Peace and Armed Conflict. (2025). *2025 Global Hunger Index: Country
rankings and severity scores* [Data set].
<https://www.globalhungerindex.org/ranking.html>

# Session info

```{r session-info}
sessionInfo()
```
